Deep Learning Models for Limnological Time Series Forecasting

### 2024-02-16 manuscript/manuscript.Rmd
Intro paragraph ideation coming redux

The time series that are used in this study are taken from the National Ecological Observatory Network's Ecological Forecasting Challenge, an open challenge where teams can submit forecasts to the publicly accessible data \cite{Quinn}. A common theme across the challenge is that a historical means (or climatology) model tends to be a top performing model \cite{Dietze, Quinn}. Thus, a primary consideration in this paper is the comparison of the machine learning models to the climatology model. If machine learning models can consistently outperform the climatology model, then machine learning is clearly a promising method class for limonological time series forecasting.

** Talk about NEON forecasting challenge and how this offers an unprecedented opportunity to test machine learning models. Typically ML models have been applied to a handful of freshwater systems; the forecasting challenge offers 34 sites across North America and Puerto Rico as a testing ground. Historically, researchers have favored using process-based models to make forecasts, but building mechanistic models for 34 sites is very challenging. Thus, leveraging non-mechanistic machine learning models offer some practical advantages. Machine learning models have been applied to limnological time series forecasting at varying levels of success. Generally, the machine learning models tended to be more basic, like RNN's and LSTM's, in the interim, new models have been developed that are said to outperform these earlier ML models. This study makes a comparison of some the newly touted state of the art machine learning models on the aquatics time series taken from NEON. From a current literature search, no previous paper has made as comprehensive of a comparison of machine learning models on the variables of water temperature, chlorophyll concentration and dissolved oxygen.

>>>>

The time series that are used in this study are taken from the National Ecological Observatory Network's Ecological Forecasting Challenge, an open challenge where teams can submit forecasts to the publicly accessible data \cite{Quinn}. A common theme across the challenge is that a historical means (or climatology) model tends to be a top performing model \cite{Dietze, Quinn}. In a model comparison for the phenology challenge, \cite{Dietze} found that the climatology outperformed all but one of the submitted models, and the best performing forecast only marginally outperformed the climatology model. Thus, a primary consideration in this paper is the comparison of the machine learning models to the climatology model. If deep learning models can consistently outperform the climatology model, then deep learning is clearly a promising method class for limonological time series forecasting.

Talk about Darts, and how there have been newly developed deep learning models that offer advantages over the more vanilla ones that are commonly used. The advantage of taking the machine learning approach is that it allows the ability to easily deploy models on systems without having a mechanistic understanding of them. This is especially advantageous in the current era where sensors can be deployed at a much faster rate than we can build accurate water quality models for every existent body of water. Like Sci-kit Learn or Tidy Models, Darts allows users to easily instantiate, fit and predict with these state of the art models without requiring users to create these models from scratch.
